---
bibliography: my_bibs.bib
header-includes:
  - \usepackage{setspace}
  - \doublespacing
  - \setlength\parindent{24pt}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \rhead{Goehring}
  - \thispagestyle{empty}
  - \usepackage[hang,flushmargin]{footmisc}
  - \makeatletter
  - \setlength{\@fptop}{0pt}
  - \makeatother
  - \usepackage[format=hang,font={normalsize,bf}]{caption}
  - \floatplacement{figure}{h!}
  - \frenchspacing

mainfont: Times New Roman
nocite: "[@hlavacStargazerWellFormattedRegression2018], [@wickhamTidyverseEasilyInstall2017],
  [@austCitrRStudioAddin2016],\n[@changExtrafontToolsUsing2014], [@rcoreteamLanguageEnvironmentStatistical2018],
  [@slowikowskiGgrepelAutomaticallyPosition2018], \n[@wickhamGtableArrangeGrobs2016],
  [@xieKnitrGeneralPurposePackage2018], [@zhuKableExtraConstructComplex2018]\n"
output:
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
  word_document: default
---
``` {r echo = FALSE, message = FALSE, results = FALSE}
library(tidyr)
library(knitr)
library(citr)
suppressMessages(library("tidyverse"))
library(ggrepel)
library(gtable)
(library(zoo))
(library(plm))
library(openxlsx)
library(readxl)
library(stats)
library(grid)
library(extrafont)

source("TANF_clean.R")
read_chunk("figures.R")

knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo = FALSE, dev = "cairo_pdf")

my_theme <- theme_classic() +
  theme(text = element_text(family = "Times New Roman"),
        panel.grid.major.y = element_line(colour = "#dedddd"),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank())
theme_set(my_theme)
```
The 1996 Personal Responsibility and Work Opportunity Reconciliation Act (PRWORA) delegated broad authority to the U.S. states over cash assistance programs for low-income families. At the legislation’s core was Temporary Assistance for Needy Families (TANF), a block grant program that provides funding to states to combat welfare dependency, support families and children, and prevent out-of-wedlock pregnancies. The creation of TANF and broad delegation of social welfare policymaking authority under the PRWORA prompt two over-arching questions. First, how have states spent TANF funds since the creation of the program? Did states use devolution to reshape welfare spending? And if changes did occur, were there broad trends among states or have states followed distinctive trajectories? Second, why do states spend TANF funds in particular ways? What factors—political, economic, or demographic—account for any observed variation in states’ TANF expenditure decisions? 

This paper addresses both questions using TANF financial data published by the Department of Health and Human Services’ Administration for Children and Families (ACF). The data describe how TANF spending evolved and diversified over time as states shifted funds away from basic assistance payments (i.e. time-limited monthly cash payments that require participants to abide by certain activity and child support requirements) toward work supports, services, and in-kind benefits. With the descriptive analysis in hand, I estimate a fixed effects regression model to examine whether state-level factors can account for the observed variation in state spending patterns between fiscal year (FY) 1998 and 2013. The analysis examines four hypotheses concerning the influence of race and ethnicity, political ideology, economic conditions, and TANF policy factors, and demonstrates that, controlling for national trends, state-level racial, ideological, and economic characteristics significantly correlate with states' cash assistance expenditures. 

### I
TANF provides each state a capped block grant and the discretion to create its own welfare program for low-income families. States' TANF block grants are neither adjusted for inflation nor, with a few minor exceptions, changes in need.^[ The PRWORA apportioned \$2 billion for a contingency fund to support states facing difficult economic conditions and, in order to further aid states during the 2009 recession, the American Recovery and Reinvestment Act allocated \$5 billion for basic assistance, emergency assistance, and employment subsidies in FY 2009 and 2010. However, the federal block grant (technically named the State Family Assistance Grant) constitutes the vast majority of federal TANF funding and is not influenced by a state's level of need [@falkTemporaryAssistanceNeedy2015].] The PRWORA apportioned states’ block grants based on the amount of federal funding received by a state for the program that preceded TANF, Aid for Families with Dependent Children (AFDC), and other low-income public assistance programs between FY 1992 and 1995; they range in size from \$21.8 million in Wyoming to $3.7 billion in California [@falkTemporaryAssistanceNeedy2015]. In addition to the federal block grant, the other main source of TANF funding is state-provided Maintenance of Effort (MOE) funds, which are set at 75% of states’ FY 1994 contributions to AFDC and other low-income public assistance programs and can increase to 80% if an insufficient number of a state’s TANF recipients are engaged in work activities [@falkTemporaryAssistanceNeedy2015].

In a particularly broad delegation of authority, the PRWORA empowers states to spend federal and MOE funds in any manner “reasonably calculated” to realize one of TANF's four statutory goals: 1) Provide assistance to needy families so that children may be cared for in their own homes or in the homes of relatives; 2) End the dependence of needy parents on government benefits by promoting job preparation, work, and marriage; 3) Prevent and reduce the incidence of out-of-wedlock pregnancies and establish annual numerical goals for preventing and reducing the incidence of these pregnancies; and 4) Encourage the formation and maintenance of two-parent families [@PublicLaw1041931996]. TANF’s statutory goals allow states to spend TANF dollars on a variety of different programs, benefits, and services, all of which can be classified as either assistance or non-assistance spending. TANF assistance includes basic assistance payments and child care and transportation benefits for families without an employed adult. TANF's time limits on benefit receipt and activity and child support requirements only pertain to recipients of TANF assistance. States may also spend federal and MOE funds on non-assistance, which includes any other spending justifiable under one of TANF's statutory goals. Non-assistance spending does not impose federal requirements on states or recipients and includes a broad array of expenditures, including refundable tax credits, work training programs, and family planning services [@falkTemporaryAssistanceNeedy2017]. 

###II
For this analysis, I group TANF expenditures into ten spending categories.^[ See Table 3 in the appendix for category groups.] Figure 1 aggregates those ten categories into three types of spending: basic assistance; work-related, in-kind, and short-term benefits; and other. In FY 1998, on average, a state spent 55.0% of total TANF expenditures on basic assistance and 19.7% on work-related, in-kind, and short-term benefits, with the remaining 26.4% dedicated to other areas, including administrative costs and transfers to other programs.^[ For more information on "other" spending, see Table 3 in the appendix, @derrUnderstandingTwoCategories2009, and @SocialServicesBlock2015.] The composition of TANF spending shifted in the years ahead as states decreased the share of TANF funds spent on basic assistance and increased proportional expenditures on work-related, in-kind, and short-term benefits. By FY 2013, the average state spent 23.6% of total TANF spending on basic assistance, a 57.1% decrease from FY 1998, and 43.2% of total TANF spending on work-related, in-kind, and short-term benefits. 

``` {r figure1, echo = FALSE, fig.align = "center", fig.cap = "Mean Proportional TANF Spending by Type, FY 1998 - 2013"}
<<Figure.1>>
```

The over two-fold increase in the average share of TANF funds dedicated to work-related, in-kind, and short-term benefits is shown in more detail in Figure 2, which displays annual proportional expenditures for the spending type's five composite categories. Over time, as states decreased basic assistance spending, they increased the share of funds allocated to a variety of different services and benefits for low-income families. Of the five categories, child care and work-related activities and supports made up the bulk of expenditures in FY 1998 and 2013. However, after a dramatic proportional increase in spending in the late 1990s, spending on both categories was roughly stagnant over the following years and remained below all-time high levels in FY 2013. In contrast, expenditures on marriage and pregnancy programs aimed at supporting healthy marriages and educating families about family planning, refundable tax credits, and diversion benefits (which usually provide one-time lump sum payments to families to help them avoid entering the state’s TANF program) consistently increased between FY 1998 and 2013. While the average state spent no TANF funds on any of the three categories in FY 1998, the categories collectively comprised 14.7% of the average state's TANF spending in FY 2013, with marriage and pregnancy programs alone comprising 7.2% 

``` {r figure2, echo = FALSE, fig.align = "center", fig.cap = "Mean Proportional Expenditures on Work-Related, In-Kind, and Short-Term Benefits, FY 1998 - 2013"}
<<Figure.2>>
```

The discussion so far has focused on the national shift away from basic assistance spending toward work supports, in-kind benefits and services. While such a focus provides an important aggregated perspective on TANF spending, the PRWORA's broad devolution of policymaking power may have impacted the degree to which a state participated in the national trend of basic assistance retrenchment. In order to more closely examine state-level trends, Figure 3 provides annual boxplots of basic assistance spending.^[ The boxplots in Figure 3 display annual median expenditures (marked by the thick black line) and the first and third quartiles (the upper and lower ends of the “box”). The lines protruding from the boxes equal the distance between the first or third quartile and the value furthest from the respective quartile that does not exceed 1.5 times the difference between the first and third quartiles. Labeled outliers are either greater or less than 1.5 times the difference between the first and third quartiles.] The boxplots show that as median basic assistance spending decreased, the distribution of state spending remained relatively constant, with annual standard deviations hovering between 10.1% (in FY 2008) and 13.8% in (in FY 1999). Consequently, even states with relatively high levels of basic assistance expenditures reduced cash assistance spending. Of the states that spent the greatest share of TANF funds on basic assistance between FY 2008 and 2013 (i.e. Maine, California, Alaska, and South Dakota), none spent more than the 75th percentile of proportional basic assistance expenditures in FY 1998 (62.6%) and only Maine exceeded the median level of basic assistance spending in FY 1998 (53.1%).

``` {r figure3, echo = FALSE, warning = FALSE, fig.align = "center", fig.cap = "Boxplots of Proportional Basic Assistance Expenditures by State, FY 1998 - 2013"}
<<Figure.3>>
```

Although the overall variation of the distribution remained relatively constant between FY 1998 and 2013, the relative rank order of states within the distribution was not static. As aggregate basic assistance spending decreased, the relative order of states was reshuffled, with many relatively higher spending states becoming relatively lower spending states and vice versa (Figure 4). For instance, of the ten states that spent the greatest portion of total TANF funds on basic assistance in FY 1998, only three – Alaska, California, and Hawaii – remained among the ten highest spending states in FY 2013. Illinois, another high-spending state in FY 1998, reduced basic assistance spending to such an extent that it was one of the ten lowest spending states in FY 2013. States that spent relatively less on basic assistance in FY 1998 also shifted their spending relative to their peers. Indiana and Oklahoma were the only states to be among the ten lowest spending states in both FY 1998 and FY 2013. On the other hand, Virginia’s relatively small decrease in proportional basic assistance spending meant it was among the ten lowest spending states in FY 1998 and the ten highest spending states in FY 2013. 

``` {r figure4, echo = FALSE, fig.align = "center", fig.cap = "Proportional Basic Assistance Expenditures in FY 1998 and 2013"}
<<Figure.4>>
```

###III 
Since the passage of the PRWORA, states have decreased the share of TANF funds dedicated to basic assistance and increased spending on work-supports, in-kind benefits, and services. However, states did not participate in this national trend in lock-step. Rather, they simultaneously participated in aggregate basic assistance retrenchment and altered their spending in distinctive ways. To better understand the variation in states’ proportional basic assistance spending, this section examines whether state-level characteristics can explain the degree to which states decreased basic assistance spending. Using a fixed effects regression model that controls for unobserved variation between states and across time, the analysis suggests that states with smaller and more racially diverse TANF caseloads, more conservative state governments, and lower unemployment rates were more likely to decrease basic assistance spending to a greater extent. 
 	
The regression analysis is grounded upon four hypotheses concerning states’ allocations of basic assistance expenditures: 1) states with more racially and ethnically diverse basic assistance caseloads spend a lower share of TANF funds on basic assistance; 2) states with more powerful and progressive democratic parties spend a greater share of TANF funds on basic assistance; 3) states with more favorable economic conditions dedicate a lower share of TANF funds to basic assistance; and 4) states’ basic assistance expenditures are sensitive to programmatic factors, such as caseload levels and work participation rates.
	
####Race and Ethnicity
Two reinforcing strands in the literature on race and social policy are especially significant when considering the potential relationships between race, ethnicity, and states’ basic assistance expenditures. The first concerns the role of racial prejudices toward African Americans in shaping public attitudes of welfare recipients. Studies such as @gilensRaceCodingWhite1996 note the significant effects of white stereotypes of African American mothers on welfare on white Americans’ support for welfare assistance. Drawing on national survey data and a randomized experiment, Gilens finds that white Americans that have significantly more negative attitudes toward African American are more likely to oppose welfare programs. Such attitudes translate to opinions of welfare policy, with “racial considerations” serving as “the single most important factor shaping whites’ views of welfare” (p. 601). 

The other strand of the literature concerns the importance of race in shaping welfare policy outcomes. Several studies have examined the correlations between race and the restrictiveness of states’ TANF policies. @sossSettingTermsRelief2001 note significant positive relationships between the respective proportions of African Americans or Latinos receiving TANF benefits in a state and the probability of a state adopting strong sanctions against participants who do not meet work requirements and a limit on the number of children that can be included in the benefit group (i.e., a “family cap”). The authors also find a significant positive relationship between the share of African Americans in a state's TANF caseload and stricter time limits on benefit receipt. @fellowesPoliticsNewAmerican2004, in another study, largely echo the conclusions of @sossSettingTermsRelief2001. They find that, on average, an increase from one standard deviation below the mean percentage of African Americans receiving TANF benefits in a state to one standard deviation above results in significantly stricter TANF benefit eligibility criteria, stricter work requirements, and lower benefit amounts. They also find that the percentage of Latinos receiving TANF benefits is consequential, with an increase from one standard deviation below the mean percentage of Latinos in a state receiving TANF benefits to one standard deviation above resulting in significantly less flexible work requirements but more lenient TANF benefit eligibility criteria.

In terms of my analysis, I hypothesize that states with greater proportions of African Americans or Hispanics in their TANF caseloads allocate a smaller share of TANF funds to basic assistance. I operationalize the hypothesis with two variables: *african americans* and *hispanics*. The former measures the percentage of adults receiving basic assistance benefits in a state who identify as African American or Black, while the latter captures the percentage of adults receiving basic assistance benefits in a state who identify as non-white and Hispanic.^[Data collected from U.S. Department of Health and Human Services, Administration for Children and Families, Office of Family Assistance: https://www.acf.hhs.gov/ofa/programs/tanf/data-reports.] 
	
####Partisan Control of State Government
Partisanship and political ideology are often considered crucial factors in structuring the scope and generosity of states’ TANF policies, with conservatives generally critical of cash welfare benefits and liberals more supportive of assistance [@romTransformingStateHealth1999]. The findings of @sossSettingTermsRelief2001 support the intuitive relationship between TANF policy and ideology. On average, the authors find that a state is 31% more likely to adopt strong sanctions for non-compliant TANF recipients if the state government is one standard deviation more conservative rather than one standard deviation more liberal. Similarly, a state is 9% more likely to adopt strong sanctions, tougher work requirements, narrower time limits, and a family cap if the state government is one standard deviation more conservative rather than one standard deviation more liberal. 

I measure state political ideology with *liberalism*. Originally developed by @berryMeasuringCitizenGovernment1998, *liberalism* captures the ideology of a state government in a calendar year from 0 (most conservative) to 100 (most liberal), weighted by the strength of the Democratic and Republican parties in the upper and lower branches of the state legislature and ideology of the governor. The authors originally measured the ideology of the political parties and governor using interest group ratings, but in @berryMeasuringCitizenGovernment2010 the authors present a slightly different measure of ideology that uses common space coordinates of Congressional roll call votes developed by @pooleRecoveringBasicSpace1998. Although they correlate strongly, @berryMeasuringCitizenGovernment2010 conclude that the updated measure of ideology is more precise than the original, and it is therefore employed for *liberalism*. 

Quantifying partisan control of state government via a measure of state government ideology controls for changes in party strength and ideology across states and time. Unlike other measures of partisanship, such as party control of state legislatures and governorships, state government ideology does not mask ideological differences between political parties in different states or shifts in political ideology over time. Instead, it incorporates these political differences and evolutions alongside swings in electoral power, creating a nuanced and flexible measure of partisanship. Following the documented relationship between support for welfare and political ideology, I hypothesize that states with higher *liberalism* scores spend a greater share of TANF funds on basic assistance. Progressive shifts in party ideology and electoral victories by Democratic candidates ought to be commensurate with increased basic assistance spending, reflecting progressive parties' general support of social welfare spending. 

####Economic Conditions
In addition to race, ethnicity, and partisan affiliation, it is important to consider the effects of state-level economic factors on basic assistance expenditures. The PRWORA was passed in an era of low unemployment, tight labor markets, and rising wages for lower-skilled workers [@blankEvaluatingWelfareReform2002]. In an extensive literature review of TANF and AFDC research in the years following the passage of the PRWORA, @blankEvaluatingWelfareReform2002 finds five econometric studies arguing that a 1% decrease in a state's unemployment rate correlates with a 5% to 7% reduction in caseload size. Since caseloads are directly related to basic assistance spending (fewer people receiving benefits allows states to spend the funds elsewhere), the studies conducted in the years following the PRWORA’s passage imply that state economic conditions ought to bear some impact on basic assistance spending. Qualitative evidence from the economic recession one decade later also supports controlling for state-level economic conditions. Thirty states saw increases in the number of basic assistance recipients following the beginning of the economic downturn in December 2007 [@zedlewskiNextStepsTemporary2010]. As economic conditions worsened, many low-income families sought cash assistance, with likely consequences for basic assistance spending. 
	
Following the established trend in the literature, I expect states with higher unemployment rates to spend greater shares of their TANF funds on basic assistance. I assess this hypothesis with *unemployment*, which measures a state’s annual unemployment rate among the civilian non-institutional population.^[ Data collected from the U.S. Department of Labor, Bureau of Labor Statistics: https://www.bls.gov/lau/rdscnp16.htm#data.] While the unemployment rate is a common metric of a state’s economic vitality, I also include *pcpi regional* as a measure of each state’s real per capita personal income in thousands of 2013 dollars controlling for regional variations in purchasing power [@blankEvaluatingWelfareReform2002].^[ Nominal values collected from the U.S. Department of Commerce, Bureau of Economic Analysis: https://www.bea.gov/regional/index.htm. Real values calculated using the Consumer Price Index for all urban consumers in, respectively, the Midwest, South, Northeast, and West regions: https://data.bls.gov/cgi-bin/surveymost?cu.] I expect per capita income to be inversely related to states’ basic assistance spending. Unemployment rates and incomes are likely to be strongly and inversely correlated, but for the population who receives TANF, they may not move in tandem. Moving from welfare to work increases earned income but may, depending on how much income a state allows TANF recipients to disregard for eligibility and benefit calculations, decrease assistance benefits. As such, controlling for states’ unemployment rates alone may not adequately control for states’ economic conditions, especially as they pertain to the low-income families most likely to receive TANF cash assistance. 

In addition to the hypothesized relationships between unemployment, per capita income, and basic assistance spending, economic conditions may also affect TANF spending by exerting fiscal pressures on states. In their study of TANF programs in California, Washington, Michigan, Florida, and Texas, @hahnStateApproachesTANF2012 note how budget deficits following the economic recession in the 2000s forced some states to reshape TANF spending. TANF’s broad spending discretion allows states to shift TANF funds away from basic assistance toward other policy areas previously funded by non-TANF dollars, allowing the latter to be used elsewhere. California, for instance, reduced basic assistance benefits by 8% in 2011 alongside other reductions in job training and child care funding, freeing $800 million in MOE expenditures for higher education programs. @hahnStateApproachesTANF2012 find evidence of a similar responses to fiscal pressures in Michigan and Washington. While shifting funds toward other purposes may help states meet budget shortfalls, the authors find that it sometimes also leads to the funding of programs obliquely related to TANF's goals. As an advocate for low-income families in Washington put it when discussing the programs being funded by TANF, “now, no one is pretending that it is for a TANF purpose” (p. 35).  

I account for a possible relationship between basic assistance spending and budgetary shortfalls with *fiscal stability*, which measures a state’s ending annual fiscal balance and budget stabilization fund (i.e. “rainy day fund”) as a percentage of its annual expenditures.^[ Data collected from the fall editions of the National Association of State Budget Officers’s *Fiscal Survey of the States*: https://www.nasbo.org/mainsite/reports-data/fiscal-survey-of-states/fiscal-survey-archives.] The hypothesized effect of *fiscal stability* on basic assistance spending is the opposite of the other economic variables included in the model, *unemployment* and *pcpi regional*. As economic conditions worsened, states experienced contradictory pressures. On the one hand, fewer jobs and lower incomes might have led states to increase basic assistance spending in order to support their residents. At the same time, worsening economic conditions reduced state revenues and strained budgets, leading to less basic assistance spending as TANF spending was reallocated to free non-TANF funds for other uses. 

####Programmatic Factors
Finally, I hypothesize that program-specific factors such as caseload sizes and work participation rates influence states' basic assistance expenditures. TANF caseloads have decreased in size since the passage of the PRWORA, a continuation of AFDC caseload reductions in the mid-1990s. Between 1998 and 2013, the number of families receiving TANF in an average month decreased by 43.4% from 3.1 to 1.7 million.^[ Data collected from U.S. Department of Health and Human Services, Administration for Children and Families, Office of Family Assistance: https://www.acf.hhs.gov/ofa/programs/tanf/data-reports.] While changes in caseload sizes are in part, as argued above, a function of economic conditions, they cannot be fully accounted for by economic explanations [@blankEvaluatingWelfareReform2002]. States are not passive actors when it comes to basic assistance eligibility and benefit levels; they control income thresholds, time limits, family cap policies, and work participation requirements – all factors which may impact how many people receive TANF assistance in a state. It is difficult to untangle the specific impact of these policies on caseload sizes and even harder to discern their relationship to basic assistance expenditures given the circularity of program policies and spending. Nevertheless, given the significant decrease in caseloads between 1998 and 2013, it is important to incorporate a broad control for changing caseloads in the model. Hence, I include *caseload*, the annual percentage change in the number of recipients in a state’s TANF caseload in an average month.^[ *caseload* includes recipients in separate state programs. Data collected from U.S. Department of Health and Human Services, Administration for Children and Families, Office of Family Assistance: https://www.acf.hhs.gov/ofa/programs/tanf/data-reports.] 

In addition to accounting for changing caseloads, it is important examine whether work participation requirements influence basic assistance spending. The PRWORA mandated that 50% of all families and 90% of two-parent families receiving TANF assistance in a state be “engaged in work” in a fiscal year in order to avoid a reduction in the state’s block grant. Before FY 2007, a state could reduce its required work participation rate by the percentage decrease in its TANF caseload from FY 1995 levels. Since caseloads declined in the years following the passage of the PRWORA, states easily met this requirement. However, the Deficit Reduction Act of 2005 made it more demanding for states to reduce their work participation rates by changing the fiscal year for calculating reductions in caseloads from FY 1995 to FY 2005.^[ The American Recovery and Reinvestment Act of 2009 suspended states' work participation rate requirements for FY 2009-2011. For more details on what constitutes being “engaged in work” and the changes to work requirement calculations see @falkTemporaryAssistanceNeedy2017.]

Since 1999 states have also been able to reduce the percentage of their caseload that must meet work requirements by spending more on MOE than mandated by federal statute. But, in addition to caseload reduction credits and excess MOE spending, states can reduce their work participation rate requirement by altering the composition of their TANF caseloads. For instance, a state can reduce the number of unemployed or difficult to employ recipients in the caseload by imposing stricter work requirements or eligibility criteria. The resulting caseload is not only more likely to meet the work participation requirement, but also be smaller and have higher earnings, resulting in lower basic assistance expenditures. 

A state can also reduce the work participation requirement by increasing the number of employed TANF recipients in the caseload. Some states have taken this approach and offer transitional benefits to TANF recipients who are ineligible due to increased earnings. For example, until October 2015, Michigan granted \$10 to former TANF recipients for 6 months after becoming ineligible due to increased earnings if they continued to meet their work requirements. Likewise, in 2016 Missouri gave one-parent families working 30 hours per week after leaving TANF \$50 for six months, and New Jersey granted former recipients $200 for 24 months if they continued to work 20 hours per week [@giannarelliWelfareRulesDatabook2017; @theurbaninstituteWelfareRulesDatabase]. The benefits allowed states to claim more employed recipients in order to satisfy the work participation rate, with the tangential effect of increasing basic assistance spending.

Since changes in the composition of the caseload in response to the work participation requirement have theoretically ambiguous effects on basic assistance spending, the influence of the work participation requirement on basic spending requires empirical testing. I evaluate the role of the work participation requirement with *wpr*, a binary variable that takes the value of one if a state did not meet its work participation rate in the prior year.^[Data collected from U.S. Department of Health and Human Services, Administration for Children and Families, Office of Family Assistance: https://www.acf.hhs.gov/ofa/programs/tanf/data-reports.] 
	
###IV 

Table 1 presents four regression models of states’ basic assistance expenditures as a percentage of total TANF expenditures.^[ In each model, the independent variables are lagged one year to correspond to the year in which the state allocated TANF funds for the next year.] Each model includes state fixed effects that control for unobserved, state-specific effects that remain constant over time. Model 1 includes all independent variables except *caseload*, *unemployment*, and *pcpi regional*. As hypothesized, the model’s racial and ethnic coefficients are highly significant and negative: A state that experiences a 1% increase in African Americans in its TANF caseload spends, on average, .727% less on basic assistance in the following fiscal year. Likewise, a state with a 1% increase in Hispanics in its TANF caseload spends .539% less on basic assistance in the following fiscal year. Model 1 also illustrates a significant relationship between whether a state met its work participation rate requirement in the prior year and its basic assistance spending. On average, a state that did not meet its work participation rate requirement spent 3.419% less on basic assistance in the following fiscal year. 

Although Model 1 displays a number of significant results, the low adjusted R^2^ value (.026) indicates that the model accounts for very little of the variation in states’ basic assistance expenditures. Models 2 and 3 improve upon Model 1 by adding *caseload*, *unemployment*, and *pcpi regional* as control variables. Model 2 accounts for the sizes of states’ TANF caseloads and demonstrates that, as hypothesized, states’ basic assistance spending is positively associated with the number of TANF cases. A percentage decrease in a state’s TANF caseload leads to a .091% average decrease in basic assistance spending in the next fiscal year. Yet, including caseload only modestly improves the explanatory power of the analysis as measured by the adjusted R^2^, which only increases from .026 in Model 1 to .041 in Model 2. 

Model 3 builds upon Model 2 by incorporating the economic variables *unemployment* and *pcpi regional*. Both variables are highly significant and negative in the model, indicating – somewhat counterintuitively – that higher unemployment and greater incomes are associated with lower proportional basic assistance spending. Controlling for economic factors also has ramifications for the significance of other variables in the model: *fiscal stability* becomes significant and positive while *hispanics* becomes insignificant. Including *unemployment* and *pcpi regional* greatly increases the portion of explained variation in basic assistance spending. Compared to Model 1, the adjusted R^2^ of Model 3 (.420) indicates that economic factors account for a sizable share of the variation in states’ spending.

``` {r include=FALSE}
suppressMessages(library("stargazer"))
```
``` {r table1, results = "asis", message = FALSE}
<<Table.1>>
```

Model 4, the final and most complete model in the analysis, introduces time fixed effects, thereby controlling for unobserved trends that remain constant across states in any given fiscal year. I argued above that there are two trends in basic assistance spending between FY 1998 and 2013. The aggregate decrease in basic assistance spending which every state participated in and the within-distribution divergence stemming from states following distinctive paths within the national trend. With time fixed effects in place, Model 4 controls for the national forces that affected basic assistance expenditures in the aggregate and allows for a more accurate assessment of the state-level characteristics that shaped the degree to which a state participated in basic assistance retrenchment. 

In Model 4, *african americans* remains highly significant and, as expected, negatively correlated with states’ basic assistance spending. On average, a state that experienced a 1% increase in the portion of its TANF caseload composed of African Americans spent .249% less on basic assistance in the following fiscal year. Such a finding corresponds to the conclusions of @gilensRaceCodingWhite1996, @fellowesPoliticsNewAmerican2004, and @sossSettingTermsRelief2001 and underlines the role that race continues to play in shaping TANF policy outcomes. In contrast, *hispanics* is neither significant nor in the hypothesized direction in the final model. The evolution of *hispanics* across the four models indicates that its significance in Models 1 and 2 was the spurious result of either omitted variable bias stemming from correlations with economic factors or national demographic changes controlled for by time fixed effects in Model 4. Regardless of the exact reason for its insignificance in the final model, *hispanics*’ positive and insignificant coefficient is not unprecedented. As mentioned above, @fellowesPoliticsNewAmerican2004 find significant inverse relationships between the percentage of Latinos receiving TANF benefits in a state and both the flexibility of work requirements and the strictness of TANF eligibility criteria. Similar to my own findings, the authors illustrate that the influence of Hispanic welfare recipients on TANF policy outcomes is not straightforward. Unlike in the case of African Americans, where there is clear evidence that negative perceptions significantly affect TANF and other social welfare policy outcomes, the share of a state's caseload composed of Hispanics seems to bear a more nuanced, undetermined influence on TANF spending. 

Turning to economic factors, Model 4 does not provide any evidence in support of my hypothesis that states with budget shortfalls will reduce basic assistance spending to cover costs. Likewise, the final model does not support my claim that states’ per capita personal incomes are negatively associated with basic assistance spending and only weakly implies that states’ unemployment rates positively correlate with basic assistance expenditures. The dramatic shifts in the magnitude and significance of *pcpi regional* and *unemployment* relative to Model 3 are a likely product of national changes in economic conditions. What appears in Model 3 as significant relationships between state-level economic variation and basic assistance spending variation are the spurious results of simultaneous aggregate movements in economic conditions and TANF spending, not potentially causal relationships at the state-level. 

Model 4 indicates that a state that experienced a 1% decline in its TANF caseload from the prior year spent, on average, .15% less on basic assistance in the following year. The increase in *caseload*’s magnitude as compared to Model 3 suggests that isolating the relationship between caseload change and basic assistance expenditures from the aggregate decreases in states’ TANF caseloads increases the direct correlation between caseload size and basic assistance spending. In other words, even when aggregate trends in caseload sizes are accounted for, states that experienced greater decreases in caseload sizes spent a lower share of their TANF block grants on basic assistance – a finding that both corresponds to my hypothesis that basic assistance spending is sensitive to caseload sizes and helps explain the observed variation within the overall trend of lower basic assistance spending. 

As discussed above, the influence of the work participation requirement on basic assistance spending is theoretically ambiguous. States can reduce the burden of the requirement by either increasing the number of employed recipients through greater basic assistance spending or decreasing the number of unemployed recipients, with the tangential effect of lower basic assistance expenditures. However, as illustrated in Model 4, the empirical relationship between the work participation requirement and basic assistance spending is clear: States that did not meet their work participation rate in the prior year spent, on average, 5.102% more on basic assistance in the following year. The highly significant coefficient rejects the hypothesis that there was a broad push among the states to remove unemployed recipients from their caseloads (and thereby create a smaller caseload with higher incomes) in response to not meeting the work participation rate requirement. 

Similar to *caseload* and *wpr*, the introduction of time fixed effects in Model 4 increases the magnitude of *liberalism*, suggesting that national changes in political ideology and aggregate changes in other state-level variables served as negative confounders in earlier models. As hypothesized, *liberalism* is positive and significant in Model 4, implying that more progressive state governments allocate larger shares of TANF funds to basic assistance. Such a finding corresponds to my hypothesis concerning progressivism and basic assistance spending as well as the well-established relationship between political ideology and social welfare spending more broadly. 

Finally, although masked in Table 1 for readability, the coefficients on the time fixed effects in Model 4 are presented below in Figure 5. The coefficients are all highly significant and of a large magnitude. Holding constant the eight state-level independent variables specified in Model 4, states spent, on average, 34.8% less on proportional basic assistance in FY 2013 than 1998. The coefficients underline the fact that although states did not decrease basic assistance spending in lock-step, they all participated in a national retrenchment of basic assistance spending. While understanding the national trend on its own, as a separate phenomenon from the role of state-level characteristics on TANF spending, is not taken up here, it certainly warrants future study. 

``` {r figure5, echo = FALSE, fig.cap = "Coefficients of Time Fixed Effects from Model 4, FY 1999 - 2013"}
<<Figure.5>>
```

###V
This paper has considered two over-arching questions: How have states spent TANF funds since the passage of the PRWORA and why do states spend TANF funds in particular ways? To answer the first question, I used TANF expenditure data to argue that since the passage of the PRWORA, states have decreased the share of TANF funds dedicated to basic assistance and increased the share spent on other forms of aid, such as child care, marriage and pregnancy programs, diversion benefits, refundable tax credits, and work-related activities and supports. While the shift away from basic assistance was sizable and widespread, with every state participating in retrenchment, I also showed that the rank order of states' basic assistance spending changed over time. The within-distribution variation suggests that state-level factors bore an impact on the degree to which a state participated in the national trend of reduced basic assistance spending.

I utilized the within-distribution variation to approach the second question of why states spent their TANF funds in certain ways. Using four hypotheses to ground my regression analysis, I find significant relationships between the share of TANF funds allocated to basic assistance and whether a state met its work participation requirement, the share of African American adults in a state's caseload, the change in a state's caseload size, and a state government's liberalism. At the same time as my findings are helpful in understanding why states took different paths within the national trend of decreasing basic assistance expenditures, they are unable to explain the national trend itself, pointing to the need for further research into the broader forces that shaped basic assistance spending. Although I have shown that a devolved program like TANF deserves a state-level analysis, I also demonstrated that national forces played a role in shaping states' funding decisions. If state-level changes require a state-level analysis, national forces require a national focus. TANF’s creation paved the way for a new approach to assisting low-income families, but also provided states with the authority to make spending decisions that have yet to be fully understood. 

\pagebreak

### APPENDIX
From FY 1997 to 2014, states reported federal TANF block grant and MOE spending to the Department of Health and Human Services via the ACF-196 form. The Office of Family Assistance (OFA), an office within the Administration for Children and Families (ACF), oversees TANF expenditure reporting and publishes annual TANF financial reports on its website.^[ https://www.acf.hhs.gov/ofa/programs/tanf/data-reports.] The published data from the ACF-196 include federal and state expenditure levels for each state and the District of Columbia across nineteen spending categories. The reporting categories on the ACF-196 did not change between FY 1997 and 2014, providing consistency in the published expenditure data. 

The use of the same reporting form and categories caters to researchers interested in TANF expenditure data, but two problems with the ACF-196’s structure complicate accurate analysis. First, the form contained broad reporting categories that were too inflexible to accurately trace changes in states’ spending over time or compare similar types of spending in different states. Without precise reporting categories, many states struggled to pair new uses for TANF dollars with available reporting categories and consequently reported spending increases in the broadly-defined “other non-assistance” and “assistance under prior law categories" [@derrUnderstandingTwoCategories2009; @johnsonTANFACFIM2013032013]. In other cases, the ACF-196 form’s reporting categories lacked clear boundaries, leading states to report similar expenditures in different categories. As the former Director of the OFA noted in regard to the ACF-196, “a state may report TANF spending for pre-school under ‘Prevention of Out-of-Wedlock Pregnancies’ or ‘Other’ and possibly even ‘Child Care,’ although the instructions specifically exclude such expenditures under child care" [@johnsonTANFACFIM2013032013].  

Analyzing TANF expenditure data is also complicated by how states reported errors. If a state discovered an error in a prior year’s expenditure report, the margin of error was subtracted or added to the respective reporting category on the current year’s ACF-196, indistinguishably blurring actual and corrected spending. The negative expenditure values in the published expenditure data are obvious evidence of this accounting method, but such cases are only the ostensible corrections where the margin of error exceeded the actual expenditures in the current year. Any value in the expenditure data can include an upward or downward correction for an error in a prior year’s report. Thus, in the words of the former Director of the OFA, it is “impossible to determine the actual TANF expenditures that occur in a fiscal year" [@johnsonTANFACFIM2013032013]. 

The flaws in the TANF expenditure data are not completely surmountable. It is impossible to know exactly where and when states misreported expenditures or corrected a prior year’s expenditure report in a later year’s report. Nevertheless, the problems can be mitigated. In order to alleviate the effects of non-mutually exclusive categories, I aggregate the nineteen distinct expenditure categories in the published data into ten using, with a few minor exceptions, the categories already developed by @schottHowStatesUse2015. As can be seen in Table 3, the aggregate categories are composed of similar ACF-196 reporting categories, reducing the probability that similar types of spending are treated as distinct in the analysis. 

In order to mitigate the effects of corrections for errors in prior year expenditure reports, I create three-year moving averages of the data.^[ The moving averages equal the three-year moving average of the quotient of basic assistance spending and total TANF expenditures. See Table 2 for regression output using raw percentages (basic assistance spending divided by total TANF expenditures, expressed as a percentage) and the percentages of the moving averages (the three-year moving average of basic assistance spending divided by the three-year moving average of total TANF expenditures, expressed as a percentage).] The three-year moving averages reduce the short-term variation in spending and prevalence of proportional expenditure values above one or below zero—the ostensible instances of states’ correcting prior years’ expenditures in the current fiscal year—from seventy-nine to fifty-six. Thus, while an improvement upon the original data, three-year moving averages do not clean all the cases of prior year corrections. Nevertheless, there is a balance to strike between clean and interesting data. Including more years in the average would capture more cases of prior year corrections, but it would also obscure actual changes in spending and inhibit longitudinal analysis. 
	
After synthesizing the original reporting categories into aggregate categories and creating three-year moving averages, my dataset includes expenditures across ten spending categories for every state and the District of Columbia from FY 1998 to 2013 expressed as percentages of total TANF expenditures, where total TANF expenditures equal federal and MOE assistance and non-assistance expenditures plus TANF funds transferred to the Social Services Block Grant and Child Care Development Fund.

``` {r table2, results = "asis", message = FALSE}
<<Table A.1>>
```

``` {r table3, echo = FALSE, results = 'asis'}
library(kableExtra)

column_1 <- c("Basic assistance", rep("Work-related, in-kind, and short-term benefits", 12), rep("Other", 6))

column_2 <- c("Basic assistance", rep("Child care", 3), rep("Work-related activities and supports", 4), rep("Refundable tax credits", 2), "Diversion benefits", rep("Marriage and pregnancy", 2), rep("Expenditures under prior law", 2), "Other non-assistance", rep("Administration and systems", 2), "Social Services Block Grant (SSBG)")

column_3 <- c("Basic assistance", "Child care (assistance)", "Child care (non-assistance)", "Transfers to the Child Care Development Fund", "Transportation and supportive services (assistance)", "Transportation (non-assistance)", "Work-related activities and expenses", "Individual development accounts", "Refundable earned income tax credits", "Other refundable tax credits", "Non-recurrent short-term benefits", "Prevention of out of wedlock pregnancies", "Two-parent family formation and maintenance", "Assistance under prior law", "Non-assistance under prior law", "Other", "Administration", "Systems", "Transfers to the Social Services Block Grant")

my_table <- tibble(column_1, column_2, column_3)

kable(my_table, "latex", align = "l", booktabs = T, caption = "TANF Spending Categories", col.names = c("Spending Types (used in Figure 1)", "Aggregate Categories", "ACF-196 Reporting Categories")) %>% 
  column_spec(1:3, width = "5cm") %>% 
  collapse_rows(columns = 1:2, latex_hline = "full", valign = "middle") %>% 
    kable_styling(font_size = 10) %>% 
  row_spec(0, bold = T)
```

\pagebreak
\setstretch{1}

# BIBLIOGRAPHY
\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\noindent
